Accelerating parameter inference with graphics processing units

### 2019-01-30 paper/paper.tex
paper.tex: minor additional profiling

\begin{table}
\begin{tabular}{lrr|cccc|r}
Version & srate & modes & $\tau_{start}$ & $\tau_{setup}$ & $\tau_{ad}$ & $\tau_{it}$ & $\frac{T_{ILE}}{N_{eval}}$\\  %\hline 
  &  & Hz & sec & sec & & $10^{-6} $sec & sec\\ \hline 
CPU & 16384 & $(2,\pm 2) $ & 20 & & & &  \editremark{900}  \\ 
       & 4096 & $(2,\pm 2) $ &   20 \\ \hline
% LIGO-WA: 
%  - old compute node: one marginalized likelihood in 2 minutes (!), 2x8
GPU (a) & 16384 & $(2,\pm 2) $  & 20  & &   \\
            & 4096 &$(2,\pm 2) $  &  20 &\\
% LIGO-CIT, pcdev13
%  - 20 iterations, about 450  to 500 sec to eval (i.e., about 25 sec per lnL)
%  -  precompute: can time manually
%   -  Overlap stages: PrecomputeLIkelihood + waveform, about 4
%  - Histogram time: about (12+30) seconds overall (histogram + ODEint)
%  - Actual likelihood evaluation: about 30 sec out of 20 eval, or about 1.5 sec
   %%   1    0.061    0.061  506.265  506.265 integrate_likelihood_extrinsic_batchmode:8(<module>)
   %%     20    0.194    0.010  498.581   24.929 integrate_likelihood_extrinsic_batchmode:690(analyze_event)
   %%     20   80.397    4.020  464.263   23.213 mcsampler.py:327(integrate)
   %%   2140   54.358    0.025  235.480    0.110 mcsampler.py:198(draw)
   %%  23530    0.134    0.000  101.280    0.004 shape_base.py:286(hstack)
   %%  24306  100.861    0.004  100.861    0.004 {numpy.concatenate}
   %%  20136    0.896    0.000   84.698    0.004 function_base.py:2063(__call__)
   %%  20136   33.608    0.002   83.802    0.004 function_base.py:2154(_vectorize_call)
   %%   2140    1.403    0.001   45.791    0.021 integrate_likelihood_extrinsic_batchmode:810(likelihood_function)
   %% 461726    0.790    0.000   36.137    0.000 polyint.py:62(__call__)
   %%   2140    5.496    0.003   30.829    0.014 factored_likelihood.py:1676(DiscreteFactoredLogLikelihoodViaArrayVectorNoLoop)
   %%    407    0.006    0.000   29.662    0.073 odepack.py:28(odeint)
   %%    407    1.360    0.003   29.644    0.073 {scipy.integrate._odepack.odeint}
   %%     20    0.052    0.003   28.785    1.439 factored_likelihood.py:92(PrecomputeLikelihoodTerms)
   %% 461726    2.809    0.000   26.983    0.000 interpolate.py:657(_evaluate)
   %%     20    0.048    0.002   24.454    1.223 lalsimutils.py:2582(hlmoft_SEOB_dict)
   %%     20    0.030    0.002   24.406    1.220 lalsimutils.py:2743(complex_hoft)
   %%     20   24.373    1.219   24.373    1.219 {lalsimulation._lalsimulation.SimInspiralChooseTDWaveform}
GPU (b) & 16384 & & 20 \\
       & 4096 & & 20 & $\simeq 1$ &   & 0.75  & 25\\ \hline
GPU (b) & 16384 & $m=\pm 2, \pm 1$ &  \\
       & 4096 & $m=\pm 2, \pm 1$ \\
\end{tabular}
% Compute node at LIGO-WA
\caption{\label{tab:CostBreakdown}Evaluation costs on default hardware, for a two-mode system $(l,m)=(2,\pm 2)$.  Scenario (a) is a compute node on LIGO-WA; scenario (b) is more modern hardware. }
\end{table}

With the new low-cost likelihood evaluations, however, the startup and setup costs  $\tau_{start},\tau_{setup}$ now can be
comparable or in excess of the total time used to evaluate the likelihood.   For this reason, we reorganize the workflow, so each
instance of ILE loops over
$N_{eval}$ different choices for $\lambda$. 
%
Additionally, particularly for massive binary black holes, we adopt two additional performance improvements.  First and
foremost, we lower the sampling rate, thus lowering the startup cost $\tau_{start}$ and particularly setup cost
$\tau_{setup}$.  Previously and out of an abundance of caution, \ILE{} employed a sampling rate $1/\Delta t = 16384
\unit{Hz}$ for all calculations, but terminated all inner products in Eq. (\ref{eq:overlap}) at roughly $f_{\rm
  max}=2048\unit{Hz}$ or less.   Reducing the sampling rate by a factor $s$ decreases the  cost of all operations with
timeseries -- they are shorter.  Depending on the relative cost of overhead versus array operations,  $\tau_{setup}$
\emph{and} the cost per evaluation $\tau_{eval}$ decrease modestly because of this effect.  
Also, to insure safe inner products over short data sets in the presence of very narrow spectral lines, we operated on a
significantly-padded data buffer,
using 
``inverse spectrum truncation'' \editremark{citation}.  For binary black holes, the degree of padding was significantly
in excess of the signal duration \editremark{numbers: say 64s for a BBH!}.  
Instead and following LI, we adopt a much shorter inverse spectrum truncation length, a much shorter padding buffer, and
a Tukey window applied to the data to remove discontinuities at the start and end of each segment.   
Reducing the duration of data analyzed by a factor $s'$ will reduce the cost $\tau_{setup}$ by a factor $s'$.  
%
Combining these factors together, the overall computational cost of each marginalized likelihood
evaluation $T_{ILE}/N_{eval}$  is
\begin{align}
T_{ILE}/N_{\rm eval} &= \frac{\tau_{start}}{N_{\rm eval}} 
 + 
 [ \tau_{setup}/s' + N_{ad} \tau_{ad} + N_{it}\tau_{it}
 ] 
\end{align}
For a typical binary black hole signal with $f_{\rm min}=20 \unit{Hz}$ and $m_1\simeq m_2\simeq 30 M_\odot$, this
version requires $\tau_{start} \simeq 12\unit{s}$, $\tau_{setup}/ss' \simeq XX$, $\tau_{ad}\simeq XXX$, and $N_{it}\tau_{\it} \simeq XX$
\editremark{numbers}.  
Because of the extremely high cost of adaptive overhead, we only adapt in two parameters, corresponding to sky location; 
and only adapt during the first iteration which returns significant $\ln {\cal L}_{\rm marg}$.  As a result, $N_{ad}$
(the number of adaptive stages \emph{per likelihood evaluation}) scales as $N_{it}/N_{chunk}/N_{\rm eval}$ and does not
increase with the overall number of samples, so the overall evaluation time scales as
\begin{align}
T_{ILE,mod}/N_{\rm eval} &= \frac{\tau_{start} + \tau_{ad}N_{it}/n_{\rm chunk}}{N_{\rm eval}} 
 \nonumber \\ &
 + 
 [ \tau_{setup}/s'  + N_{it}\tau_{it}
 ] 
\end{align}
Typically we target $N_{\rm eval} \simeq 20$ so that startup and other one-time costs  are not a significant contributor to the overall
runtime.   In this configuration, the run time per marginalized likelihood is around $T_{ILE}/N_{\rm eval} \simeq 25
\unit{s}$ for a binary black hole \editremark{checkme - old number}.    Of this $25 \unit{s}$, only roughly
$1.5\unit{s}$ corresponds to evaluating the likleihood, implying the cost $\tau_{it} \simeq 1.5/N_{\rm it} \simeq
7.5\times 10^{-7}$..

@@

The \ILE{} likelhood  generates waveform dynamics and $h_{lm}(t)$ once per evaluation $\lambda$.   Waveform generation can contribute significantly to the
time needed to evaluate  $\ln {\cal L}_{\rm marg}$ for extremely costly
waveform models which require $\tau_{wf} $ a significant fraction of $  \tau_{setup}$.   That said, the tests described
above used relatively costly waveform generation with $h_{lm}(t)$ allocation requiring $\tau_{wf} \simeq 1 \unit{s}$.

>>>>

\begin{table}
\begin{tabular}{lrr|ccccc|r}
Version & srate & modes & $\tau_{start}$ & $\tau_{setup}$ & $\tau_{ad}$ & $\tau_{it,like}$ &$\tau_{it,rest}$ & $\frac{T_{ILE}}{N_{eval}}$\\  %\hline 
  &  & Hz & sec & sec & & $\mu$sec & $\mu$sec\\ \hline 
CPU & 16384 & $(2,\pm 2) $ & 20 & && & &  \editremark{900}  \\ 
       & 4096 & $(2,\pm 2) $ &   20 \\ \hline
% LIGO-WA: 
%  - old compute node: one marginalized likelihood in 2 minutes (!), 2x8
GPU (a) & 16384 & $(2,\pm 2) $  & 20  & &   \\
            & 4096 &$(2,\pm 2) $  &  20 &\\
%% % LIGo CIT, pcdev13, 16 kHz : basically indistinguishable from cost at 4kHz
%%       1    0.027    0.027  560.572  560.572 integrate_likelihood_extrinsic_batchmode:8(<module>)
%%        20    0.082    0.004  552.004   27.600 integrate_likelihood_extrinsic_batchmode:690(analyze_event)
%%        20   76.306    3.815  503.342   25.167 mcsampler.py:327(integrate)
%%      2345   57.019    0.024  241.951    0.103 mcsampler.py:198(draw)
%%     25785    0.135    0.000   99.535    0.004 shape_base.py:286(hstack)
%%     26561   99.102    0.004   99.102    0.004 {numpy.concatenate}
%%     21981    0.933    0.000   87.819    0.004 function_base.py:2063(__call__)
%%     21981   34.750    0.002   86.887    0.004 function_base.py:2154(_vectorize_call)
%%      2345    1.391    0.001   86.327    0.037 integrate_likelihood_extrinsic_batchmode:810(likelihood_function)
%%      2345    0.007    0.000   46.648    0.020 __init__.py:630(asnumpy)
%%      2345   46.557    0.020   46.640    0.020 {method 'get' of 'cupy.core.core.ndarray' objects}
%%        20    0.179    0.009   43.145    2.157 factored_likelihood.py:92(PrecomputeLikelihoodTerms)
%%        20    0.174    0.009   37.498    1.875 lalsimutils.py:2582(hlmoft_SEOB_dict)
%%        20    0.099    0.005   37.324    1.866 lalsimutils.py:2743(complex_hoft)
%%        20   37.220    1.861   37.220    1.861 {lalsimulation._lalsimulation.SimInspiralChooseTDWaveform}
%%    444365    0.761    0.000   35.303    0.000 polyint.py:62(__call__)
%%      2345    5.260    0.002   33.956    0.014 factored_likelihood.py:1676(DiscreteFactoredLogLikelihoodViaArrayVectorNoLoop)
%%       407    0.005    0.000   28.757    0.071 odepack.py:28(odeint)
%%       407    1.327    0.003   28.740    0.071 {scipy.integrate._odepack.odeint}
%%    444365    2.644    0.000   26.423    0.000 interpolate.py:657(_evaluate)
%% 117270874   22.569    0.000   22.569    0.000 mcsampler.py:685(uniform_samp)
% LIGO-CIT, pcdev13, 4kHz
%  - 20 iterations, about 450  to 500 sec to eval (i.e., about 25 sec per lnL)
%  -  precompute: can time manually
%   -  Overlap stages: PrecomputeLIkelihood + waveform, about 4
%  - Histogram time: about (12+30) seconds overall (histogram + ODEint)
%  - Actual likelihood evaluation: about 30 sec out of 20 eval, or about 1.5 sec
   %%   1    0.061    0.061  506.265  506.265 integrate_likelihood_extrinsic_batchmode:8(<module>)
   %%     20    0.194    0.010  498.581   24.929 integrate_likelihood_extrinsic_batchmode:690(analyze_event)
   %%     20   80.397    4.020  464.263   23.213 mcsampler.py:327(integrate)
   %%   2140   54.358    0.025  235.480    0.110 mcsampler.py:198(draw)
   %%  23530    0.134    0.000  101.280    0.004 shape_base.py:286(hstack)
   %%  24306  100.861    0.004  100.861    0.004 {numpy.concatenate}
   %%  20136    0.896    0.000   84.698    0.004 function_base.py:2063(__call__)
   %%  20136   33.608    0.002   83.802    0.004 function_base.py:2154(_vectorize_call)
   %%   2140    1.403    0.001   45.791    0.021 integrate_likelihood_extrinsic_batchmode:810(likelihood_function)
   %% 461726    0.790    0.000   36.137    0.000 polyint.py:62(__call__)
   %%   2140    5.496    0.003   30.829    0.014 factored_likelihood.py:1676(DiscreteFactoredLogLikelihoodViaArrayVectorNoLoop)
   %%    407    0.006    0.000   29.662    0.073 odepack.py:28(odeint)
   %%    407    1.360    0.003   29.644    0.073 {scipy.integrate._odepack.odeint}
   %%     20    0.052    0.003   28.785    1.439 factored_likelihood.py:92(PrecomputeLikelihoodTerms)
   %% 461726    2.809    0.000   26.983    0.000 interpolate.py:657(_evaluate)
   %%     20    0.048    0.002   24.454    1.223 lalsimutils.py:2582(hlmoft_SEOB_dict)
   %%     20    0.030    0.002   24.406    1.220 lalsimutils.py:2743(complex_hoft)
   %%     20   24.373    1.219   24.373    1.219 {lalsimulation._lalsimulation.SimInspiralChooseTDWaveform}
GPU (b) & 16384 & & 20  & 1.8 & 1 & 0.85& 12 &28\\
       & 4096 & & 20 & $1.2 $ &  1  & 0.75 &  & 25\\ \hline
GPU (b) & 16384 & $\pm 2, \pm 1$ &  \\
       & 4096 & $\pm 2, \pm 1$ \\
\end{tabular}
% Compute node at LIGO-WA
\caption{\label{tab:CostBreakdown}Evaluation costs on default hardware, for a two-mode system $(l,m)=(2,\pm 2)$.  Scenario (a) is a compute node on LIGO-WA; scenario (b) is more modern hardware. }
\end{table}

With the new low-cost likelihood evaluations, however, the startup and setup costs  $\tau_{start},\tau_{setup}$ now can be
comparable or in excess of the total time used to evaluate the likelihood.   In fact, the total time per Monte Carlo
evaluation $\tau_{it}=\tau_{manage}+\tau_{eval}$  spent in general-purpose overhead $\tau_{manage}$ will be much larger than the time $\tau_{eval}$ spent
carrying out scientific calculations by evaluationg the likelihood.  For these reasons, we reorganize the workflow, so each
instance of ILE loops over
$N_{eval}$ different choices for $\lambda$. 
%
Additionally, particularly for massive binary black holes, we investigate two additional performance improvements.  First and
foremost, we lower the sampling rate, thus lowering the startup cost $\tau_{start}$ and particularly setup cost
$\tau_{setup}$.  Previously and out of an abundance of caution, \ILE{} employed a sampling rate $1/\Delta t = 16384
\unit{Hz}$ for all calculations, but terminated all inner products in Eq. (\ref{eq:overlap}) at roughly $f_{\rm
  max}=2048\unit{Hz}$ or less.   Reducing the sampling rate by a factor $s$ will the  cost of all operations with
timeseries -- they are shorter.  Depending on the relative cost of overhead versus array operations,  $\tau_{setup}$
\emph{and} the cost per evaluation $\tau_{eval}$ decrease modestly because of this effect.  
Also, to insure safe inner products over short data sets in the presence of very narrow spectral lines, we operated on a
significantly-padded data buffer,
using 
``inverse spectrum truncation'' \editremark{citation}.  For binary black holes, the degree of padding was significantly
in excess of the signal duration \editremark{numbers: say 64s for a BBH!}.  
Instead and following LI, we adopt a much shorter inverse spectrum truncation length, a much shorter padding buffer, and
a Tukey window applied to the data to remove discontinuities at the start and end of each segment.   
Reducing the duration of data analyzed by a factor $s'$ will reduce the cost $\tau_{setup}$ by a factor $s'$.  
%
Combining these factors together, the overall computational cost of each marginalized likelihood
evaluation $T_{ILE}/N_{eval}$  is
\begin{align}
T_{ILE}/N_{\rm eval} &= \frac{\tau_{start}}{N_{\rm eval}} 
 + 
 [ \tau_{setup}/s' + N_{ad} \tau_{ad} + N_{it}\tau_{it}
 ] 
\end{align}
For a typical binary black hole signal with $f_{\rm min}=20 \unit{Hz}$ and $m_1\simeq m_2\simeq 30 M_\odot$, this
version requires $\tau_{start} \simeq 12\unit{s}$, $\tau_{setup}/ss' \simeq XX$, $\tau_{ad}\simeq XXX$, and $N_{it}\tau_{\it} \simeq XX$
\editremark{numbers}.  
Because of the extremely high cost of adaptive overhead, we only adapt in two parameters, corresponding to sky location; 
and only adapt during the first iteration which returns significant $\ln {\cal L}_{\rm marg}$.  As a result, $N_{ad}$
(the number of adaptive stages \emph{per likelihood evaluation}) scales as $N_{it}/N_{chunk}/N_{\rm eval}$ and does not
increase with the overall number of samples, so the overall evaluation time scales as
\begin{align}
T_{ILE,mod}/N_{\rm eval} &= \frac{\tau_{start} + \tau_{ad}N_{it}/n_{\rm chunk}}{N_{\rm eval}} 
 \nonumber \\ &
 + 
 [ \tau_{setup}/s'  + N_{it}\tau_{it}
 ] 
\end{align}
Typically we target $N_{\rm eval} \simeq 20$ so that startup and other one-time costs  are not a significant contributor to the overall
runtime.   In this configuration, the run time per marginalized likelihood is around $T_{ILE}/N_{\rm eval} \simeq 25
\unit{s}$ for a binary black hole \editremark{checkme - old number}.    Of this $25 \unit{s}$, only roughly
$1.5\unit{s}$ corresponds to evaluating the likleihood, implying the cost $\tau_{it} \simeq 1.5/N_{\rm it} \simeq
7.5\times 10^{-7}$.
Likewise, because overhead dominates over array manipulations, an analysis with  $16\unit{kHz}$ timeseries requires only
marginally more time than the corresponding $4\unit{kHz}$ analysis.

@@

The \ILE{} likelhood  generates waveform dynamics and $h_{lm}(t)$ once per evaluation $\lambda$.   Waveform generation can contribute significantly to the
time needed to evaluate  $\ln {\cal L}_{\rm marg}$ for extremely costly
waveform models which require $\tau_{wf} $ a significant fraction of $  \tau_{setup}$.   That said, the tests described
above used relatively costly waveform generation with $h_{lm}(t)$ allocation requiring $\tau_{wf} \simeq 1.2 \unit{s}$
at 4kHz and $1.9\unit{s}$ at 16 kHz.
